---
title: "School analysis using multilevel models"
author: "Daniel Berry and George Rooney"
date: "Oct 11 2016"
output:
  revealjs::revealjs_presentation:
    self_contained: true
---

# Introduction 
* For this project we sought to model mathematics pass rate in schools
using a combination 
```{r, echo = F}
library(plyr)
library(reshape2)
library(stringr)
library(mice)
set.seed(123456)
```

# Loading the data into R

## Load in initial data 
```{r}
subject_data <- read.csv('2015-16_school_subject.csv', stringsAsFactors = FALSE)

accred_data <- read.csv('accreditation_2013_and_after_report.csv', stringsAsFactors = FALSE, skip = 3)

ethnicity_data <- read.csv('school_summaries_ethnicity.csv', stringsAsFactors = FALSE, skip = 4)
ethnicity_data$School.No. <- gsub('\\,','',ethnicity_data$School.No.)
ethnicity_data$School.No.[ethnicity_data$School.No. == 260] <- 231

salary_data <- read.csv('salaries.csv',stringsAsFactors = FALSE)

vote_data <- read.csv('2012_general_results.csv', stringsAsFactors = FALSE)

truancy_data <- read.csv('truancy.csv', stringsAsFactors = FALSE)
truancy_data$X <- NULL
truancy_data$X.1 <- NULL
truancy_data <- truancy_data[1:123,]
```

## Load in 2: more loading
```{r, warning=FALSE}
subject_data_high <- subset(subject_data, High.Grade == 12)

ethnicity_data_high <- subset(ethnicity_data, Grade == '12')

vote_data_pres <- subset(vote_data, OfficeTitle == 'President and Vice President' & Party %in% c('Democratic','Republican'))
vote_temp <- melt(vote_data_pres, id.vars = c('LocalityName', 'Party'), measure.vars = 'TOTAL_VOTES')
vote_cast <- dcast(vote_temp, formula = LocalityName + Party ~ variable, fun.aggregate = sum)

vote_cast_2 <- dcast(vote_cast, LocalityName ~ Party, value.var = 'TOTAL_VOTES')
vote_cast_3 <- cbind(vote_cast_2, vote_cast_2[,2:3]/rowSums(vote_cast_2[,2:3]))
names(vote_cast_3) <- c(names(vote_cast_2), paste(names(vote_cast_2[,2:3]), '.pct', sep=''))

salary_data$FY.2014..Actual.Average.Teacher.Salary <- as.numeric(gsub('\\t|\\s|\\,', '', salary_data$FY.2014..Actual.Average.Teacher.Salary))

salary_data$FY.2015..Actual.Average.Teacher.Salary <- as.numeric(gsub('\\t|\\s|\\,', '', salary_data$FY.2015..Actual.Average.Teacher.Salary))

salary_data$FY.2016..Budgeted.Average.Teacher.Salary <- as.numeric(gsub('\\t|\\s|\\,', '', salary_data$FY.2016..Budgeted.Average.Teacher.Salary))
```
# Aggregating the data

## Merge subject dataset with accredidation dataset:
```{r}
all_data <- merge(x = subject_data_high,
                  y = accred_data,
                  by.x = c('Div.Num','Sch.Num'),
                  by.y = c('Division.Number','School.Number'),
                  suffixes = c('.sdh','.acd'),
                  all = TRUE)
```


## Merge on ethnicity data:
```{r}
all_data <- merge(x = all_data,
                  y = ethnicity_data_high,
                  by.x = c('Div.Num', 'Sch.Num'),
                  by.y = c('Division.No.', 'School.No.'),
                  suffixes = c('.ad', '.ed'),
                  all = TRUE)
```

## merge on truancy data: 
```{r}
all_data <- merge(x = all_data,
                  y = truancy_data,
                  by.x = c('Div.Num'),
                  by.y = c('Division.No'),
                  suffixes = c('.ad','.t'),
                  all = TRUE)
```

## merge on salary data
```{r}
all_data <- merge(x = all_data,
                  y = salary_data,
                  by.x = 'Div.Num',
                  by.y = 'Division',
                  suffixes = c('.ad', 'sd'),
                  all = TRUE)
```

## merge on voting data: 
```{r}
all_data$Div.Name <- tolower(str_trim(all_data$Div.Name))
vote_cast_3$LocalityName <- tolower(vote_cast_3$LocalityName)

all_data <- merge(x = all_data,
                  y = vote_cast_3,
                  by.x = 'Div.Name',
                  by.y = 'LocalityName',
                  suffixes = c('.ad','.vc2'),
                  all = TRUE)
```

## compute truancy as a percentage: 
```{r, echo = FALSE, warning = FALSE}
to_num_cols <- c(grep('Male|Female', names(all_data), value = TRUE), 'English', 'Science', 'Mathematics', 'History', 'Total..Full.time...Part.time.Students', 'Truancy.Count', 'GCI', grep('Pass', names(all_data), value = TRUE))

for (col in to_num_cols) {
    all_data[[col]] <- as.numeric(all_data[[col]])
}

to_pct_cols <- grep('Male|Female', names(all_data), value = TRUE)
for (col in to_pct_cols) {
    all_data[[paste0(col,'.pct')]] <- all_data[[col]]/all_data$Total..Full.time...Part.time.Students
}
```

```{r}
## compute truancy percentage:
truancy_data_pct <- ddply(all_data,
                          .(Div.Name),
                          function(df) {
                              mean(df$Truancy.Count) / sum(df$Total..Full.time...Part.time.Students)
                          }
                          )

names(truancy_data_pct) <- c('Div.Name', 'Truancy.pct')

all_data <- merge(x = all_data,
                  y = truancy_data_pct,
                  by = 'Div.Name',
                  all = TRUE)
```

## convert ethnicity to percent of school composition:
```{r}
races <- c('Native.Hawaiian', 'Asian','White', 'Two.or.more.races', 'Hispanic', 'American.Indian', 'Black')
for (race in races) {
    race_cols <- grep('.pct',grep(race, names(all_data), value = TRUE), value = TRUE)
    all_data[[paste0(race,'.pct')]] <- rowSums(all_data[,race_cols])
}
```

# Examining the modeling dataset

## Selecting the modeling dataset
```{r}
model_data <- subset(all_data,
                     Subgroup == 'All Students' & Subject == 'Mathematics',
                     select = c('Div.Name',
                                'Div.Num',
                                'School.Name',
                                'Sch.Num',
                                paste0(races,'.pct'),
                                grep('Pass', names(all_data), value = TRUE),
                                'School.Accreditation.Rating',
                                grep('English|Science|Mathematics|History', names(all_data), value = TRUE),
                                'Total..Full.time...Part.time.Students',
                                grep('Salary', names(all_data), value = TRUE),
                                'Democratic.pct',
                                'Truancy.pct'
                                ))
```



# Exploring the data: 

## Some plots:
The following are some univariate plots of the predictor variables in the model (using complete cases only)

## Plot of 'Asian.pct':
```{r, out.width = 600}
ggplot(model_data, aes_string(x = 'Asian.pct', y = 'Mathematics')) + geom_point() + geom_smooth(method = 'lm')
```

## Plot of 'Hispanic.pct':
```{r, out.width = 600}
ggplot(model_data, aes_string(x = 'Hispanic.pct', y = 'Mathematics')) + geom_point() + geom_smooth(method = 'lm')
```

## Plot of 'Black.pct':
```{r, out.width = 600}
ggplot(model_data, aes_string(x = 'Black.pct', y = 'Mathematics')) + geom_point() + geom_smooth(method = 'lm')
```

## Plot of 'X2014.2015.Pass.Rate':
```{r, out.width = 600}
ggplot(model_data, aes_string(x = 'X2014.2015.Pass.Rate', y = 'Mathematics')) + geom_point() + geom_smooth(method = 'lm')
```

## Plot of 'English':
```{r, out.width = 600}
ggplot(model_data, aes_string(x = 'English', y = 'Mathematics')) + geom_point() + geom_smooth(method = 'lm')
```

## Plot of 'History':
```{r, out.width = 600}
ggplot(model_data, aes_string(x = 'History', y = 'Mathematics')) + geom_point() + geom_smooth(method = 'lm')
```

## Plot of 'Science':
```{r, out.width = 600}
ggplot(model_data, aes_string(x = 'Science', y = 'Mathematics')) + geom_point() + geom_smooth(method = 'lm')
```

## Plot of 'Total..Full.time Students':
```{r, out.width = 600}
ggplot(model_data, aes_string(x = 'Total..Full.time...Part.time.Students', y = 'Mathematics')) + geom_point() + geom_smooth(method = 'lm')
```

## Plot of 'FY.2016..Budgeted.Salary': 

```{r, out.width = 600}
ggplot(model_data, aes_string(x = 'FY.2016..Budgeted.Average.Teacher.Salary', y = 'Mathematics')) + geom_point() + geom_smooth(method = 'lm')
```

## Some collearity:
```{r}
car::vif(lm(Mathematics ~ . -Div.Name -School.Name -Sch.Num -Div.Num, model_data))
```

## Simplify the dataset: 
```{r}
model_data$Native.Hawaiian.pct <- NULL
model_data$American.Indian.pct <- NULL
model_data$Two.or.more.races.pct <- NULL
model_data$White.pct <- NULL

model_data$FY.2014..Actual.Average.Teacher.Salary <- NULL
model_data$FY.2015..Actual.Average.Teacher.Salary <- NULL

model_data$X2013.2014.Pass.Rate <- NULL
model_data$X2015.2016.Pass.Rate <- NULL

model_data$Met.English <- NULL
model_data$Met.History <- NULL
model_data$Met.Science <- NULL

model_data$Met.Mathematics[model_data$Met.Mathematics == 'Not Tested'] <- NA
```

## Which leaves final correlations:
```{r}
knitr::kable(cor(model_data[,sapply(model_data, is.numeric)]), digits = 2)
```

## Unfortunately there's some missing data: 
```{r}
print(paste('Number missing: ',sum(!complete.cases(model_data))))
```

## Missing data analysis

For 2 categorical variables with missing levels: created a new level for missing:

```{r}
model_data$Met.Mathematics[is.na(model_data$Met.Mathematics)] <- "Missing"
model_data$School.Accreditation.Rating[is.na(model_data$School.Accreditation.Rating)] <- "Missing"
```

For numeric variables we did imputation via `mice`:

```{r}
mice_hs <- mice(model_data)
model_data_mice <- complete(mice_hs)
```

## Split into train/test:
```{r}
split <- rbinom(nrow(model_data_mice),1,.8)
model_data_T <- model_data_mice[split*(1:length(split)),]
model_data_V <- model_data_mice[-(split*(1:length(split))),]
```

# First model: complete pooling

## Build the model
Model: $$Math_i = \beta_0 + \sum_{j=1}^{17} \beta_i\cdot x_{ji}$$
```{r}
summary(complete_T <- lm(Mathematics ~ ., data = model_data_T[,-c(1:4)]))
```

## Examine coefficients:
```{r, echo = FALSE}
pander::pander(summary(complete_T))
```

## Predict on new data:
```{r}
cpooling_pred <- predict.lm(complete_T, newdata = model_data_V[,-c(1:4)])
(cp_mse <- sum((model_data_V$Mathematics-cpooling_pred)^2))
```

## Predicted vs fitted:
```{r, out.width = 600}
plot(model_data_V$Mathematics, cpooling_pred)
```

## Predicted vs residual
```{r, out.width = 600}
plot(model_data_V$Mathematics, (model_data_V$Mathematics-cpooling_pred))
```

# No pooling model

## Issues:
Due to the fact that there are many districts with only one observation in them, we can't just simply fit an no pooling model, too many coefficients. However, we should be able to place priors on the coefficients and do a bayesian hierarchical model. So far we've only got this working for random slopes for 1 covariate, not sure why we can't do three. 

## Fit the model
```{r}
no_pooling_2_T <- blmer(Mathematics ~ Asian.pct +
                        Hispanic.pct +
                        Black.pct +
                        X2014.2015.Pass.Rate +
                        School.Accreditation.Rating +
                        English +
                        Met.Mathematics +
                        History +
                        Science +
                        Total..Full.time...Part.time.Students +
                        FY.2016..Budgeted.Average.Teacher.Salary +
                        (1 + FY.2016..Budgeted.Average.Teacher.Salary | Div.Num),
                      model_data_T,
                      cov.prior = wishart)
```

## Predict on new data
```{r, out.width = 600}
npooling_pred <- predict(no_pooling_2_T, newdata = model_data_V, allow.new.levels=T)
(np_mse <- sum((model_data_V$Mathematics-npooling_pred)^2))
```

## Predicted fitted plot
```{r, out.width = 600}
plot(model_data_V$Mathematics, npooling_pred)
```

## Predicted vs residual
```{r, out.width = 600}
plot(model_data_V$Mathematics, (model_data_V$Mathematics-npooling_pred))
```

# Partial pooling models: random intercept 

## Random slope model specification:
$$Math_i = \alpha_{J[i]} + \sum_{j=1}^{17}\beta_j*X_{ji}$$

## Fit model:
```{r}
partial_pooling_T <- lmer(Mathematics ~ Asian.pct +
                          Hispanic.pct +
                          Black.pct +
                          X2014.2015.Pass.Rate +
                          School.Accreditation.Rating +
                          English +
                          Met.Mathematics +
                          History +
                          Science +
                          Total..Full.time...Part.time.Students +
                          FY.2016..Budgeted.Average.Teacher.Salary +
                          Democratic.pct +
                          Truancy.pct +
                          (1 | Div.Num),
                          model_data_T)
```

## Examine fit:
```{r}
summary(partial_pooling_T)
```

## Predict on new data
```{r}
ppooling_pred <- predict(partial_pooling_T, newdata = model_data_V, allow.new.levels=T)
(pp_mse <- sum((model_data_V$Mathematics-ppooling_pred)^2))
```

## Predicted fitted plot
```{r, out.width = 600}
plot(model_data_V$Mathematics, ppooling_pred)
```

## Predicted residual plot
```{r, out.width = 600}
plot(model_data_V$Mathematics, (model_data_V$Mathematics-ppooling_pred))
```

## Some code to make caterpillar plots
inspired by: http://stackoverflow.com/questions/11123147/dotplot-of-random-effects
```{r}
caterpillar_plot <- function(model) {
    theRan <- ranef(model, condVar=TRUE)
    pv <- attr(theRan$Div.Num, "postVar")
    se <- pv[1, 1, ]
    theIntercepts <- theRan$Div.Num[, 1, drop=F]
    theFrame <- cbind(theIntercepts, se)
    names(theFrame)[1] <- "Intercept"
    theFrame$Low <- with(theFrame, Intercept - 2 * se)
    theFrame$High <- with(theFrame, Intercept + 2 * se)
    theFrame$Variable <- rownames(theFrame)
    freqs <- lapply(names(ranef(model)), function(x) cbind(ranef(model)[[x]], table(model.frame(model)[[x]])))[[1]]
    theFrame <- merge(x = theFrame, y = freqs, by.x = c('Intercept','Variable'), by.y = c('(Intercept)','Var1'))
    p <- ggplot(theFrame, aes(y=Intercept, x=Freq)) + geom_linerange(aes(ymin=Low, ymax=High), colour="black") + geom_point(, colour="blue")  + labs(y="Random Intercept", x = 'Number of Schools in District',title='Estimate +- SE')
    return(p)
}
```

## View caterpillar plot:
```{r, out.width = 600}
pp_plot <- caterpillar_plot(partial_pooling_T)
plot(pp_plot)
```



# Final results

## Model comparison:

Partial pooling performed the best in terms of held out MSE:
```{r}
cp_mse
np_mse
pp_mse
```

## Takeaway:
Using the partial pooling model, we have that school's mathematics passing score is best predicted by its overall pass rate last year, it's accredidation rating, percentage Hispanic, and size. 

## Thoughts:
- Perhaps using other priors in the no pooling model would have worked better?
